Functional pant indicators - wetland, indicator documentation

### 2023-05-03 DRAFT_functional plant indicators_wetland.Rmd
Update DRAFT_functional plant indicators_wetland.Rmd
corrected some typos and formatting issues directly in the file

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture on would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). This plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

## About the underlying data
In the 'plant functional indicator' project for wetlands, We use three sets of data for building indicators for ecological condition
- as testing data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. Within a 5-year cycle, 1000 sites are  randomly distributed across mainland Norway. Each ANO site spans a 500 x 500 m grid cell and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as on of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) NiN reference data:
The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1 > 1/32, 2 = 1/32 - 1/8 , 3 = 1/8 - 3/8, 4 = 3/8 - 4/5, 5 = 3/8 - 4/5 + dominance, 6 = 4/5 -1 ). For the purpose of this project these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(3) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub-/tree encroachment, drainage, and pollution.

### Representativity in time and space
For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country. As wetlands occur more often in certain regions of Norway than in others the amount of plots and sites is not equal among the five regions ecological condition is assessed for (Northern Norway, Central Norway, Eastern Norway, Western Norway, Southern Norway). The in total 1351 plots are distributed across regions in the following way:

Northern Norway: 416
Central Norway:  362
Eastern Norway:  295
Western Norway:  151
Southern Norway: 127

### Temporal coverage
The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. Thus there is no actual time series to these data and the indicator evaluation does therefore not include any temporal analyses.

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al. 2018). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for ecosystem sub-types (grunntyper or kartleggingsenheter at a 1:5000 mapping scale) within the main wetland types in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (non-parametric here) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH) we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson 2021.

@@

Median = reference values
0.025 and 0.975 quantiles = lower and upper limit values
min and max of the respective indicator's scale = min/max values

@@

Once testing data (ANO), and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO-plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type. An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

>>>>

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al. 2019) for wetland ecosystems. We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen. These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021).

## About the underlying data
In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:

- as test data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
- Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) NiN reference data:
The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5). For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(3) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub/tree encroachment, drainage, and pollution.

### Representativity in time and space
For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country. As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 1351 plots are distributed across regions in the following way:

- Northern Norway: 416
- Central Norway:  362
- Eastern Norway:  295
- Western Norway:  151
- Southern Norway: 127

### Temporal coverage
The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

@@

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al. 2018). By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

@@

- Median = reference values
- 0.025 and 0.975 quantiles = lower and upper limit values
- min and max of the respective indicator's scale = min/max values

@@

Once test data (ANO) and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type. An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.
